Measuring the double-drift illusion with hand trajectories

### 2021-12-16 Manuscript.Rmd
last few changes demanded by reviewers

---
title: "Measuring the double-drift illusion with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document:
    reference_docx: template.docx
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$. That is, the deviation from the physical direction is given by:

@@

In particular, we compare the illusion strengths we find with those reported by Heller et al. [-@Heller2021]. They show a double-drift stimulus with 36 combinations of internal and external motion for 500 ms and then have participants indicate on a ring where they perceived the path of the stimulus to intersect with this ring. While they hypothesize that illusion strength decreases with faster external motion (perhaps because of resets or limits on perception), here we will use their overall estimate of illusion strength, given by $K=0.74$. While in _dva/s_ our stimuli have relatively high external and relatively low internal motion, compared to Heller et al. [-@Heller2021], they are within the range of tested stimuli as is the predicted illusion strength.

@@

```{r fig3, echo=F, fig.height=3.4125, fig.width=6, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets. **Left:** At some point, the perceived position may stop moving further away from its actual path. For some people this takes the form of a ‘return’ reset, back toward the true position either suddenly or slowly, for others it means the illusory position remains at a fixed offset moving parallel to the real path. **Right:** Given varying strength of the illusion (denoted by the angle of the purple arrows), e.g. due to stimulus properties or within or between subject differences, resets may occur at different points in space. **Top right:** if these resets can be explained by a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate; the spatial offset from the vertical path. **Bottom right:** if they occur instead after some period of time, resets should have a constant radius (R); the time since the gabor’s motion began."}
library(magick)
Fig3 <- image_read_pdf('doc/methods_fig_2.pdf')
par(mai=c(0,0,0,0))
plot(Fig3)
```

@@

For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported themsleves as right handed and had either normal or corrected-to-normal vision. Procedures were in accordance with the Declaration of Helsinki (2003) and were approved by York’s Human Participants Review Committee. All participants provided prior, written, informed consent.

@@

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or to the time from stimulus onset, but instead may be caused by a more "random" process.

@@

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories from Liu et al. [-@Liu2021], re-analyzed here. **A:** The Gabor’s internal motion was adjusted for each condition and participant so that the perceived path was always locally straight down (yellow arrows). The 350 ms temporal gap then ensured a full reset [@Lisi2015] to the physical location when the Gabor continued on its path following the gap (blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (the vertical green lines) indicated the amount of illusion left at the end of  the pre-gap path (measured PSE). **B:** Percent reduction of the illusion (difference between measured offset and the offset expected with no intervening spontaneous resets  as a percentage of expected offset) as a function of path length and duration. The size of the illusion decreased over time for a fixed path length and local illusion angle. These data are consistent with resets occurring randomly over time at ~0.7 / _s_ modeled as a Poisson process."}
plotJumpPSEs(target='inline', individualDots=F, modelDots=F, gammaDots=F, modelLine=T, gammaLine=T, propExpect=T)
```

>>>>

---
title: "Measuring the double-drift illusion with hand trajectories"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document:
    reference_docx: template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{amsmath}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al. [-@Heller2021]. A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but we are interested in illusion strength only, so we use the simple vector combination model. In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$. That is, the deviation from the physical direction is given by:

@@

In particular, we compare the illusion strengths we find with those reported by Heller et al. [-@Heller2021]. They show a double-drift stimulus with 36 combinations of internal and external motion for 500 ms and then have participants indicate on a ring where they perceived the path of the stimulus to intersect with this ring. While they hypothesize that illusion strength decreases with faster external motion (perhaps because of resets or limits on perception), here we will use their overall estimate of illusion strength, given by $K=0.74$. While in _dva/s_ our stimuli have relatively high external and relatively low internal motion, compared to Heller et al. [-@Heller2021], they are within the range of tested stimuli as are the predicted illusion strengths.

@@

```{r fig3, echo=F, fig.height=3.4125, fig.width=6, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets. **Left:** At some point, the perceived position may stop moving further away from its actual path. For some people this takes the form of a ‘return’ reset, back toward the true position either suddenly or slowly, for others it means the illusory position remains at a fixed offset moving parallel to the real path. **Right:** Given varying strength of the illusion (denoted by the angle of the purple arrows), e.g. due to stimulus properties or within or between subject differences, resets may occur at different points in space. **Top right:** if these resets can be explained by a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate; the spatial offset from the vertical path. **Bottom right:** if they occur instead after some period of time, resets should have a constant radius (T); the time since the gabor’s motion began."}
library(magick)
Fig3 <- image_read_pdf('doc/methods_fig_2.pdf')
par(mai=c(0,0,0,0))
plot(Fig3)
```

@@

For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported themselves as right handed and had either normal or corrected-to-normal vision. Procedures were in accordance with the Declaration of Helsinki (2003) and were approved by York’s Human Participants Review Committee. All participants provided prior, written, informed consent.

@@

In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type. We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. Neither of these provided a viable explanation of the pattern of resets. Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or to the time from stimulus onset, but instead may be caused by a more "random" process. Overall, both the spatial offsets and time of resets seem to follow a gamma distribution about equally well (Fig 5A, marginal distributions). However, average spatial offsets seems to be lower with lower Y coordinates, such that the gamma distribution on spatial offsets doesn’t work as well throughout the data set (Fig 5C). In contrast, a gamma distribution on reset time works equally well, irrespective of angle (Fig 5B), and this is reflected in the better fit of this model. Hence it seems most likely that reset points are randomly distributed over time, according to a gamma distribution.

@@

```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories from Liu et al. [-@Liu2021], re-analyzed here. **A:** The Gabor’s internal motion was adjusted for each condition and participant so that the perceived path was always locally straight down (yellow arrows). The 350 ms temporal gap then ensured a full reset [@Lisi2015] to the physical location when the Gabor continued on its path following the gap (blue arrow). Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (the vertical green lines) indicated the amount of illusion left at the end of the pre-gap path (measured illusion), which can be contrasted with the expected illusion if there were no spontaneous resets. **B:** Percent reduction of the illusion (difference between measured illusion and the expected illusion with no intervening spontaneous resets as a percentage of expected illusion) as a function of path length and duration. The size of the illusion decreased over time for a fixed path length and local illusion angle. These data are consistent with resets occurring randomly over time at ~0.7 / _s_ modeled as a Poisson process."}
plotJumpPSEs(target='inline', individualDots=F, modelDots=F, gammaDots=F, modelLine=T, gammaLine=T, propExpect=T)
```
